Informing policy via dynamic models: Cholera in Haiti

### 2023-08-15 ms-plos.Rnw
changed run_level
minor edits + updated TODO list

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Title of submission to PLOS journals} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jesse Wheeler\textsuperscript{1*},
AnnaElaine L. Rosengart\textsuperscript{2},
Zhuoxun Jiang\textsuperscript{1},
Kevin Tan\textsuperscript{3},
Noah Treutle\textsuperscript{1},
Edward L. Ionides\textsuperscript{1}
\\
\bigskip
\textbf{1} Statistics Department, University of Michigan, Ann Arbor, Michigan, USA
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

@@

For any time point in $t_{1:N}$, the process model describes the movement of individuals from one latent state compartment to another.
Per-capita transition rates are given in Eqs.~\myeqref{model1:SE}-\myeqref{model1:birth}:
\begin{eqnarray}
\label{model1:SE}
\mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t),
\\%[-3pt]
\label{model1:EI}
\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\vaccineEfficacy_\vaccCounter(t)\big),
\\%[-3pt]
\label{model1:EA}
\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \vaccineEfficacy_\vaccCounter(t),
\\%[-3pt]
\label{model1:toR}
\mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR,
\\%[-3pt]
\label{model1:RS}
\mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS,
\\%[-3pt]
\label{model1:vacc}
\mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t),
\\%[-3pt]
\label{model1:death}
\mu_{S_{\vaccCounter}\demography} &=& \mu_{E_{\vaccCounter}\demography} = \mu_{I_{\vaccCounter}\demography} = \mu_{A_{\vaccCounter}\demography}=\mu_{R_{\vaccCounter}\demography} = \delta,
\\
\label{model1:birth}
\mu_{\demography S_0} &=& \muBirth,
\end{eqnarray}
where $\vaccCounter\in \seq{0}{\vaccClass}$.
Here, $\mu_{AB}$ is a transition rate from compartment $A$ to $B$.
We have an additional demographic source and sink compartment $\demography$ modeling entry into the study population due to birth or immigration, and exit from the study population due to death or immigration.
Thus, $\mu_{A\demography}$ is a rate of exiting the study population from compartment $A$ and $\mu_{\demography B}$ is a rate of entering the study population into compartment $B$.

@@

The latent state vector $\bm{X}(t)$ is initialized by setting each compartment count for each vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
The measurement model assumes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
See Sec.~S3 of the supplement material for more details.

@@

Individuals can progress to a latent infection $E_{u\vaccCounter}$ followed by symptomatic infection $I_{u\vaccCounter}$ with recovery to $R_{u\vaccCounter}$ or asymptomatic infection $A_{u\vaccCounter}$ with recovery to $R^A_{u\vaccCounter}$.
The force of infection depends on both direct transmission and an aquatic reservoir, $W_u(t)$, and is given by
\begin{equation}
\label{model2:lambda}
\lambda_{u}(t) = 0.5\big(1+\seasAmplitude \cos(2\pi t + \phaseParm)\big)
\frac{\beta_W\, W_u(t)}{ \Wsat  + W_u(t)} +
\transmission \left\{\sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\}.
\end{equation}
The latent state is therefore described by the vector $\bm{X}(t) = \big(S_{u\vaccCounter}(t),\allowbreak E_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}^A(t),\allowbreak W_u,\allowbreak u \in \seq{1}{U},\allowbreak \vaccCounter \in \seq{0}{4}\big)$.
The cosine term in Eq.~\myeqref{model2:lambda} accounts for annual seasonality, with a phase parameter $\phaseParm$.
The Lee et al.~\cite{lee20} implementation of Model~2 fixes $\phaseParm = 0$.

@@

Per-capita transition rates are given in Eqs.~\myeqref{eq:model3:SI}--\myeqref{eq:model3:Decay}.
\begin{eqnarray}
\label{eq:model3:SI}
\mu_{S_{u\vaccCounter}I_{u}} &=& \symptomFrac \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
\\%[-3pt]
\label{eq:model3:SA}
\mu_{S_{u\vaccCounter}A_{u}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
\\%[-3pt]
\label{eq:model3:IR}
\mu_{I_{u}R_{u\vaccCounter 1}} &=& \mu_{A_{u}R_{u\vaccCounter 1}} = \muIR,
\\%[-3pt]
\label{eq:model3:IS}
\mu_{I_{u}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u}S_{u0}} = \muDeath
\\%[-3pt]
\label{eq:model3:RRnext}
\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
\\%[-3pt]
\label{eq:model3:RS}
\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
\\%[-3pt]
\label{eq:model3:water}
\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
\\%[-3pt]
\label{eq:model3:Decay}
\mu_{W_u\demography} &=& \Wremoval.
\end{eqnarray}
As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in Eqs.~\myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
In Eq.~\myeqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $\mathrm{Den}_u$ is the population density.
Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in Eq.~\myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
All deaths are balanced by births into the susceptible compartment in Eqs.~\myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.

@@

We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \cite{ionides17}.
Lee et al.~\cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
These results are suggestive that the inclusion of a trend in transmission rate improves the quantitative ability of Model~1 to describe the observed data.
The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.

@@

We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
One change we made in the statistical analysis that led to larger model likelihoods was increasing the computational effort in the numerical maximization.
While IF2 enables parameter estimation for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations.
Many Monte Carlo replications are then required to quantify and further reduce the error.
The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Informing policy via dynamic models: Cholera in Haiti} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jesse Wheeler\textsuperscript{1*},
AnnaElaine L. Rosengart\textsuperscript{2},
Zhuoxun Jiang\textsuperscript{1},
Kevin Tan\textsuperscript{3},
Noah Treutle\textsuperscript{1},
Edward L. Ionides\textsuperscript{1}
\\
\bigskip
\textbf{1} Statistics Department, University of Michigan, Ann Arbor, Michigan, USA
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

@@

For any time point in $t_{1:N}$, the process model $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ is defined by the describing how individuals move from one latent state compartment to another.
Per-capita transition rates are given in Eqs.~\myeqref{model1:SE}-\myeqref{model1:birth}:
\begin{eqnarray}
\label{model1:SE}
\mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t),
\\%[-3pt]
\label{model1:EI}
\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\vaccineEfficacy_\vaccCounter(t)\big),
\\%[-3pt]
\label{model1:EA}
\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \vaccineEfficacy_\vaccCounter(t),
\\%[-3pt]
\label{model1:toR}
\mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR,
\\%[-3pt]
\label{model1:RS}
\mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS,
\\%[-3pt]
\label{model1:vacc}
\mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t),
\\%[-3pt]
\label{model1:death}
\mu_{S_{\vaccCounter}\demography} &=& \mu_{E_{\vaccCounter}\demography} = \mu_{I_{\vaccCounter}\demography} = \mu_{A_{\vaccCounter}\demography}=\mu_{R_{\vaccCounter}\demography} = \delta,
\\
\label{model1:birth}
\mu_{\demography S_0} &=& \muBirth,
\end{eqnarray}
where $\vaccCounter\in \seq{0}{\vaccClass}$.
Here, $\mu_{AB}$ is a transition rate from compartment $A$ to $B$.
We have an additional demographic source and sink compartment $\demography$ modeling entry into the study population due to birth or immigration, and exit from the study population due to death or immigration.
Thus, $\mu_{A\demography}$ is a rate of exiting the study population from compartment $A$ and $\mu_{\demography B}$ is a rate of entering the study population into compartment $B$.

@@

The latent state vector $\bm{X}(t)$ is initialized by setting the counts for each compartment and vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
The measurement model assumes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
See Sec.~S3 of the supplement material for more details.

@@

Like Model~1, the process model $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ is primarily defined via the description of movement of individuals between compartments, however Model~2 also includes a dynamic description of a latent bacterial compartment as well.
Individuals can progress to a latent infection $E_{u\vaccCounter}$ followed by symptomatic infection $I_{u\vaccCounter}$ with recovery to $R_{u\vaccCounter}$ or asymptomatic infection $A_{u\vaccCounter}$ with recovery to $R^A_{u\vaccCounter}$.
The force of infection depends on both direct transmission and an aquatic reservoir, $W_u(t)$, and is given by
\begin{equation}
\label{model2:lambda}
\lambda_{u}(t) = 0.5\big(1+\seasAmplitude \cos(2\pi t + \phaseParm)\big)
\frac{\beta_W\, W_u(t)}{ \Wsat  + W_u(t)} +
\transmission \left\{\sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\}.
\end{equation}
The latent state is therefore described by the vector $\bm{X}(t) = \big(S_{u\vaccCounter}(t),\allowbreak E_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}^A(t),\allowbreak W_u,\allowbreak u \in \seq{1}{U},\allowbreak \vaccCounter \in \seq{0}{4}\big)$.
The cosine term in Eq.~\myeqref{model2:lambda} accounts for annual seasonality, with a phase parameter $\phaseParm$.
The Lee et al.~\cite{lee20} implementation of Model~2 fixes $\phaseParm = 0$.

@@

Per-capita transition rates are used for both compartments representing human counts and the aquatic reservoir of bacteria; these rates  are given in Eqs.~\myeqref{eq:model3:SI}--\myeqref{eq:model3:Decay}.
\begin{eqnarray}
\label{eq:model3:SI}
\mu_{S_{u\vaccCounter}I_{u}} &=& \symptomFrac \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
\\%[-3pt]
\label{eq:model3:SA}
\mu_{S_{u\vaccCounter}A_{u}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
\\%[-3pt]
\label{eq:model3:IR}
\mu_{I_{u}R_{u\vaccCounter 1}} &=& \mu_{A_{u}R_{u\vaccCounter 1}} = \muIR,
\\%[-3pt]
\label{eq:model3:IS}
\mu_{I_{u}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u}S_{u0}} = \muDeath
\\%[-3pt]
\label{eq:model3:RRnext}
\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
\\%[-3pt]
\label{eq:model3:RS}
\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
\\%[-3pt]
\label{eq:model3:water}
\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
\\%[-3pt]
\label{eq:model3:Decay}
\mu_{W_u\demography} &=& \Wremoval.
\end{eqnarray}
As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in Eqs.~\myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
In Eq.~\myeqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $\mathrm{Den}_u$ is the population density.
Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in Eq.~\myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
All deaths are balanced by births into the susceptible compartment in Eqs.~\myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.

@@

We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \cite{ionides17}.
Lee et al.~\cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.

@@

We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
One change we made in the statistical analysis that led to larger model likelihoods was increasing the computational effort in the numerical maximization.
While IF2 enables parameter estimation for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations.
Many Monte Carlo replications are then required to quantify and further reduce the error.
The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
